Component response rate variation underlies the stability of complex systems

### 2020-03-20 notebook/ms.tex
Fix the notation in several places, particularly around the diagonal matrix gamma versus the vector diag(gamma)

\setlength{\droptitle}{-2em}

  \title{Component response rate variation underlies the stability of complex
systems}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{A. Bradley Duthie (
\href{mailto:alexander.duthie@stir.ac.uk}{\nolinkurl{alexander.duthie@stir.ac.uk}}
)}
    \preauthor{\centering\large\emph}
  \postauthor{\par}
      \predate{\centering\large\emph}
  \postdate{\par}
    \date{Biological and Environmental Sciences, University of Stirling, Stirling,
UK, FK9 4LA}

@@

\textbf{Component response rates of random complex systems}. Complex
systems (\(\mathbf{M}\)) are built from two matrices, one modelling
component interactions (\(\mathbf{A}\)), and second modelling component
response rates (\(\gamma\)). Both \(\mathbf{A}\) and \(\gamma\) are
square \(S \times S\) matrices. Rows in \(\mathbf{A}\) define how a
given component \(i\) is affected by each component \(j\) in the system,
including itself (where \(i = j\)). Off-diagonal elements of
\(\mathbf{A}\) are independent and identically distributed (i.i.d), and
diagonal elements are set to \(A_{ii} = -1\) as in
May\textsuperscript{\protect\hyperlink{ref-May1972}{1}}. Diagonal
elements of \(\gamma\) are positive, and off-diagonal elements are set
to zero (i.e. \(\gamma\) is a diagonal matrix with positive support).
The distribution of \(diag(\gamma)\) over \(S\) components thereby
models the distribution of component response rates. The dynamics of the
entire system \(\mathbf{M}\) can be defined as
follows\textsuperscript{\protect\hyperlink{ref-Patel2018}{19}},

\begin{equation} \label{defM}
M = \gamma A.
\end{equation}

Equation \ref{defM} thereby serves as a null model to investigate how
variation in component response rate (\(\sigma^{2}_{\gamma}\)) affects
complex systems. In the absence of such variation
(\(\sigma^{2}_{\gamma} = 0\)), \(\gamma\) is set to the identity matrix
(diagonal elements all equal 1) and \(\mathbf{M} = \mathbf{A}\). Under
these conditions, eigenvalues of \(\mathbf{M}\) are distributed
uniformly\textsuperscript{\protect\hyperlink{ref-Tao2010}{14}} in a
circle centred at \((-d, 0)\) with a radius of
\(\sigma \sqrt{SC}\)\textsuperscript{\protect\hyperlink{ref-May1972}{1}}
(Fig. 1a).

@@

The increase in \(\sigma^{2}_{M}\) caused by \(\sigma^{2}_\gamma\) can
be visualised from the eigenvalue spectra of \(\textbf{A}\) versus
\(\textbf{M} = \gamma\textbf{A}\) (Fig. 1). Given \(d = 0\) and
\(C = 1\), the distribution of eigenvalues of \(\textbf{A}\) and
\(\textbf{M}\) lie within a circle of a radius \(\sigma_{A}\sqrt{S}\)
and \(\sigma_{M}\sqrt{S}\), respectively (Fig. 1a vs.~1b). If
\(d \neq 0\), positive \(\sigma^{2}_\gamma\) changes the distribution of
eigenvalues\textsuperscript{\protect\hyperlink{ref-Ahmadian2015}{21}--\protect\hyperlink{ref-Stone2017}{23}},
potentially affecting stability (Fig. 1c vs.~1d).

@@

If \(\rho < 0\), such as when \(\textbf{M}\) models a predator-prey
system in which \(M_{ij}\) and \(M_{ji}\) have opposing signs, stability
increases\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}}. If
diagonal elements of \(\gamma\) vary independently, the magnitude of
\(\rho\) is decreased because \(\sigma^{2}_{\gamma}\) increases the
variance of \(M_{ij}\) without affecting the expected covariance between
\(M_{ij}\) and \(M_{ji}\) (Figure 2).

@@

\textbf{Targeted manipulation of \(\mathbf{\gamma}\)}. To further
investigate the potential of \(\sigma^{2}_{\gamma}\) to be stabilising,
I used a genetic algorithm. Genetic algorithms are heuristic tools that
mimic evolution by natural selection, and are useful when the space of
potential solutions (in this case, possible combinations of \(\gamma\)
values leading to stability in a complex system) is too large to search
exhaustively\textsuperscript{\protect\hyperlink{ref-Hamblin2013}{29}}.
Generations of selection on \(\gamma\) value combinations to minimise
\(\Re(\lambda_{max})\) demonstrated the potential for
\(\sigma^{2}_{\gamma}\) to increase system stability. Across
\(S = \{2, 3, ..., 39, 40\}\), sets of \(\gamma\) values were found that
resulted in stable systems with probabilities that were up to four
orders of magnitude higher than when \(\gamma = 1\) (see Supplementary
Information), meaning that stability could often be achieved by
manipulating \(S\) \(\gamma\) values rather than \(S \times S\)
\(\mathbf{M}\) elements (i.e., by manipulating component response rates
rather than interactions between components).

@@

My focus here is distinct from Gibbs et
al.\textsuperscript{\protect\hyperlink{ref-Gibbs2017}{22}}, who applied
the same mathematical framework to investigate how a diagonal matrix
\(\mathbf{X}\) (equivalent to \(\gamma\) in my model) affects the
stability of a community matrix \(\mathbf{M}\) given an interaction
matrix \(\mathbf{A}\) within a generalised Lotka-Volterra model, where
\(\mathbf{M} = \mathbf{XA}\). Gibbs et
al.\textsuperscript{\protect\hyperlink{ref-Gibbs2017}{22}} analytically
demonstrated that the effect of \(\mathbf{X}\) on system stability
decreases exponentially as system size becomes arbitrarily large
(\(S \to \infty\)) for a given magnitude of complexity
\(\sigma\sqrt{SC}\). My numerical results do not contradict this
prediction because I did not scale \(\sigma = 1 / \sqrt{S}\), but
instead fixed \(\sigma\) and increased \(S\) to thereby increase total
system complexity (see Supplemental Information for results simulated
across \(\sigma\) and \(C\)). Overall, I show that component response
rate variation increases the upper bound of complexity at which
stability can be realistically observed, meaning that highly complex
systems are more likely than not to vary in their component response
rates, and for this variation to be critical for system stability.

@@

A general mathematical framework encompassing shifts in eigenvalue
distributions caused by a diagona matrix \(\gamma\) has been
investigated\textsuperscript{\protect\hyperlink{ref-Ahmadian2015}{21}}
and recently applied to questions concerning species density and
feasibility\textsuperscript{\protect\hyperlink{ref-Gibbs2017}{22},\protect\hyperlink{ref-Stone2017}{23}},
but \(\gamma\) has not been interpreted as rates of response of
individual system components to perturbation. My model focuses on
component response rates for systems of a finite size, in which
complexity is high but not yet high enough to make the probability of
stability unrealistically low for actual empirical systems. For this
upper range of system size, randomly assembled complex systems are more
likely to be stable if their component response rates vary (e.g.,
\(10 < S < 30\) for parameter values in Fig. 4). Variation in component
response rate might therefore be critical for maintaining stability in
many highly complex empirical systems. These results are broadly
applicable for understanding the stability of complex networks across
the physical, life, and social sciences.

@@

\textbf{Component response rate (\(\mathbf{\gamma}\)) variation}. In a
synthesis of eco-evolutionary feedbacks on community stability, Patel et
al. model a system that includes a vector of potentially changing
species densities (\(\mathbf{n}\)) and a vector of potentially evolving
traits
(\(\mathbf{x}\))\textsuperscript{\protect\hyperlink{ref-Patel2018}{19}}.
For any species \(i\) or trait \(j\), change in species density
(\(n_{i}\)) or trait value (\(x_{j}\)) with time (\(t\)) is a function
of the vectors \(\mathbf{n}\) and \(\mathbf{x}\),

@@

\[\frac{d\mathbf{v}(t)}{dt} = \mathbf{\gamma} \mathbf{A}\mathbf{v}(t).\]

In the above, \(\mathbf{\gamma}\) is a diagonal matrix in which elements
correspond to individual component response rates. Therefore,
\(\mathbf{M} = \mathbf{\gamma} \mathbf{A}\) defines the change in values
of system components and can be analysed using the techniques of
May\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Ahmadian2015}{21},\protect\hyperlink{ref-May1973}{30}}.
In these analyses, row means of \(\mathbf{A}\) are expected to be
identical, but variation around this expectation will naturally arise
due to random sampling of \(\mathbf{A}\) off-diagonal elements and
finite \(S\). In simulations, the total variation in \(\mathbf{M}\) row
means that is attributable to \(\mathbf{A}\) is small relative to that
attributable to \(\mathbf{\gamma}\), especially at high \(S\). Variation
in \(\mathbf{\gamma}\) specifically isolates the effects of differing
component response rates, hence causing differences in expected
\(\mathbf{M}\) row means.

\textbf{Construction of random and structured networks}. I used the R
programming language for all numerical simulations and
analyses\textsuperscript{\protect\hyperlink{ref-Rproject}{31}}. Purely
random networks were generated by sampling off-diagonal elements from an
i.i.d \(A_{ij} \sim \mathcal{N}(0, 0.4)\) with a probability \(C\)
(unsampled elements were set to zero). Diagonal elements \(A_{ii}\) were
set to \(-1\). Elements of \(\gamma\) were simulated i.i.d. from a
distribution with positive support (typically
\(\gamma_{i} \sim \mathcal{U}(0, 2)\)). Random \(\mathbf{A}\) matrices
with correlated elements \(A_{ij}\) and \(A_{ji}\) were built using
Cholesky decomposition. Competitor networks in which off-diagonal
elements \(A_{ij} \geq 0\) were constructed by first building a random
\(\mathbf{A}\), then flipping the sign of any elements in which
\(A_{ij} < 0\). Similarly, mutualist networks were constructed by
building a random \(\mathbf{A}\), then flipping the sign of elements
where \(A_{ij} > 0\). Predator-prey networks were constructed by first
buing a random \(\mathbf{A}\), then flipping the sign of either
\(A_{ij}\) or \(A_{ji}\) if \(A_{ij} \times A_{ji} > 0\).

@@

Cascade food-webs were constructed following Solow and
Beet\textsuperscript{\protect\hyperlink{ref-Solow1998}{17}}. First, a
random matrix \(\mathbf{A}\) was generated with off-diagonal elements
sampled i.i.d so that \(A_{ij} \sim \mathcal{N}(0, 0.4)\). Each
component in the system was ranked from \(1\) to \(S\). If component
\(i\) had a higher rank than component \(j\) and \(A_{ij} < 0\), then
\(A_{ij}\) was multiplied by \(-1\). If \(i\) had a lower rank than
\(j\) and \(A_{ji} < 0\), then \(A_{ji}\) was multiplied by \(-1\). In
practice, this resulted in a matrix \(\mathbf{A}\) with negative and
positive values in the lower and upper triangles, respectively. Diagonal
elements of \(\mathbf{A}\) were set to \(-1\) and \(C = 1\). I simulated
cascade food webs across all combinations of
\(S = \{2, 3, ..., 59, 60\}\).

@@

\textbf{Genetic algorithm}. Ideally, to investigate the potential of
\(\sigma^{2}_{\gamma}\) for increasing the proportion of stable complex
systems, the search space of all possible \(\gamma\) vectors would be
evaluated for each unique \(\mathbf{M = \gamma A}\). This is technically
impossible because \(\gamma_{i}\) can take any real value between 0-2,
but even rounding \(\gamma_{i}\) to reasonable values would result in a
search space too large to practically explore. Under these conditions,
genetic algorithms are highly useful tools for finding practical
solutions by mimicking the process of biological
evolution\textsuperscript{\protect\hyperlink{ref-Hamblin2013}{29}}. In
this case, the practical solution is finding vectors of
\(\mathbf{\gamma}\) that decrease the most positive real eigenvalue of
\(\mathbf{M}\). The genetic algorithm used achieves this by initialising
a large population of 1000 different potential \(\mathbf{\gamma}\)
vectors and allowing this population to evolve through a process of
mutation, crossover (swaping \(\gamma_{i}\) values between vectors),
selection, and reproduction until either a \(\mathbf{\gamma}\) vector is
found where all \(\Re(\lambda) < 0\) or some ``giving up'' critiera is
met.

For each \(S = \{2, 3, ..., 39, 40\}\), the genetic algorithm was run
for 100000 random \(\mathbf{M = \gamma A}\) (\(\sigma_{A} = 0.4\),
\(C = 1\)). The genetic algorithm was initialised with a population of
1000 different \(\mathbf{\gamma}\) vectors with elements sampled i.i.d
from \(\gamma_{i} \sim \mathcal{U}(0, 2)\). Eigenanalysis was performed
on the \(\mathbf{M}\) resulting from each \(\mathbf{\gamma}\) vector,
and the 20 \(\mathbf{\gamma}\) vectors resulting in \(\mathbf{M}\) with
the lowest \(\Re(\lambda_{max})\) each produced 50 clonal offspring with
subsequent random mutation and crossover between the resulting new
generation of 1000 \(\mathbf{\gamma}\) vectors. Mutation of each
\(\gamma_{i}\) in a \(\mathbf{\gamma}\) vector occurred with a
probability of 0.2, resulting in a mutation effect of size
\(\mathcal{N}(0, 0.02)\) being added to generate the newly mutated
\(\gamma_{i}\) (any \(\gamma_{i}\) values that mutated below zero were
multiplied by \(-1\), and any values that mutated above 2 were set to
2). Crossover occurred between two sets of 100 \(\mathbf{\gamma}\)
vectors paired in each generation; vectors were randomly sampled with
replacement among but not within sets. Vector pairs selected for
crossover swapped all elements between and including two \(\gamma_{i}\)
randomly selected with replacement (this allowed for reversal of vector
element positions during crossover; e.g.,
\(\{\gamma_{4}, \gamma_{5}, \gamma_{6}, \gamma_{7}\} \to \{\gamma_{7}, \gamma_{6}, \gamma_{5}, \gamma_{4}\}\)
). The genetic algorithm terminated if a stable \(\mathbf{M}\) was
found, 20 generations occurred, or if the mean \(\mathbf{\gamma}\)
fitness increase between generations was less than 0.01 (where fitness
was defined as \(W_{\gamma} = -\Re(\lambda_{max})\) for \(\mathbf{M}\)).

@@

\textbf{Figure 1: Eigenvalue distributions of random complex systems.}
Each panel shows the real (x-axis) and imaginary (y-axis) parts of
\(S =\) 400 eigenvalues from random \(S \times S\) matrices.
(\(\textbf{a}\)) A system represented by a matrix \(\mathbf{A}\), in
which all elements are sampled from a normal distribution with
\(\mu = 0\) and \(\sigma_{A} = 1/\sqrt{S}\). Points are uniformly
distributed within the blue circle centred at the origin with a radius
of \(\sigma_{A} \sqrt{S} = 1\). (\(\textbf{b}\)) The same system as
\(\textbf{a}\) after including variation in the response rates of \(S\)
components, represented by the diagonal matrix \(\gamma\), such that
\(\mathbf{M} = \gamma\mathbf{A}\). Elements of \(\gamma\) are randomly
sampled from a uniform distribution from \(\min = 0\) to \(\max = 2\).
Eigenvalues of \(\mathbf{M}\) are then distributed non-uniformly within
the red circle centred at the origin with a radius of
\(\sqrt{\sigma^{2}_{A}(1 + \sigma^{2}_{\gamma})S} \approx\) 1.11.
(\(\textbf{c}\)) A different random system \(\mathbf{A}\) constructed
from the same parameters as in \(\textbf{a}\), except with diagonal
values of \(-1\). (\(\textbf{d}\)) The same system \(\textbf{c}\) after
including variation in component response rates, sampled from
\(\mathcal{U}(0, 2)\) as in \(\textbf{b}\).

@@

\textbf{Figure 4: Stability of large complex systems with and without
variation in component response rate (\(\boldsymbol{\gamma}\)).} The
\(\ln\) number of systems that are stable across different system sizes
(\(S = \{2, 3, ..., 49, 50 \}\)) given \(C = 1\), and the proportion of
systems for which variation in \(\gamma\) is critical for system
stability. For each \(S\), 1 million complex systems are randomly
generated. Stability of each complex system is tested given variation in
\(\gamma\) by randomly sampling \(\gamma \sim \mathcal{U}(0, 2)\).
Stability given \(\sigma^{2}_{\gamma}>0\) is then compared to stability
in an otherwise identical system in which
\(\gamma = E[\mathcal{U}(0, 2)]\) for all components. Blue and red bars
show the number of stable systems in the absence and presence of
\(\sigma^{2}_{\gamma}\), respectively. The black line shows the
proportion of systems that are stable when \(\sigma^{2}_{\gamma}>0\),
but would be unstable if \(\sigma^{2}_{\gamma}=0\).

>>>>

\setlength{\droptitle}{-2em}
  \title{Component response rate variation underlies the stability of complex
systems}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{A. Bradley Duthie (
\href{mailto:alexander.duthie@stir.ac.uk}{\nolinkurl{alexander.duthie@stir.ac.uk}}
)}
  \preauthor{\centering\large\emph}
  \postauthor{\par}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{Biological and Environmental Sciences, University of Stirling, Stirling,
UK, FK9 4LA}

@@

\textbf{Component response rates of random complex systems}. Complex
systems (\(\mathbf{M}\)) are built from two matrices, one modelling
component interactions (\(\mathbf{A}\)), and second modelling component
response rates (\(\boldsymbol{\gamma}\)). Both \(\mathbf{A}\) and
\(\boldsymbol{\gamma}\) are square \(S \times S\) matrices. Rows in
\(\mathbf{A}\) define how a given component \(i\) is affected by each
component \(j\) in the system, including itself (where \(i = j\)).
Off-diagonal elements of \(\mathbf{A}\) are independent and identically
distributed (i.i.d), and diagonal elements are set to \(A_{ii} = -1\) as
in May\textsuperscript{\protect\hyperlink{ref-May1972}{1}}. Diagonal
elements of \(\boldsymbol{\gamma}\) are positive, and off-diagonal
elements are set to zero (i.e. \(\boldsymbol{\gamma}\) is a diagonal
matrix with positive support). The distribution of
\(diag(\boldsymbol{\gamma})\) over \(S\) components thereby models the
distribution of component response rates. The dynamics of the entire
system \(\mathbf{M}\) can be defined as
follows\textsuperscript{\protect\hyperlink{ref-Patel2018}{19}},

\begin{equation} \label{defM}
\mathbf{M} = \boldsymbol{\gamma} \mathbf{A}.
\end{equation}

Equation \ref{defM} thereby serves as a null model to investigate how
variation in component response rate (\(\sigma^{2}_{\gamma}\)) affects
complex systems. In the absence of such variation
(\(\sigma^{2}_{\gamma} = 0\)), \(\boldsymbol{\gamma}\) is set to the
identity matrix (diagonal elements all equal 1) and
\(\mathbf{M} = \mathbf{A}\). Under these conditions, eigenvalues of
\(\mathbf{M}\) are distributed
uniformly\textsuperscript{\protect\hyperlink{ref-Tao2010}{14}} in a
circle centred at \((-d, 0)\) with a radius of
\(\sigma \sqrt{SC}\)\textsuperscript{\protect\hyperlink{ref-May1972}{1}}
(Fig. 1a).

@@

The increase in \(\sigma^{2}_{M}\) caused by \(\sigma^{2}_\gamma\) can
be visualised from the eigenvalue spectra of \(\textbf{A}\) versus
\(\textbf{M} = \boldsymbol{\gamma}\textbf{A}\) (Fig. 1). Given \(d = 0\)
and \(C = 1\), the distribution of eigenvalues of \(\textbf{A}\) and
\(\textbf{M}\) lie within a circle of a radius \(\sigma_{A}\sqrt{S}\)
and \(\sigma_{M}\sqrt{S}\), respectively (Fig. 1a vs.~1b). If
\(d \neq 0\), positive \(\sigma^{2}_\gamma\) changes the distribution of
eigenvalues\textsuperscript{\protect\hyperlink{ref-Ahmadian2015}{21}--\protect\hyperlink{ref-Stone2017}{23}},
potentially affecting stability (Fig. 1c vs.~1d).

@@

If \(\rho < 0\), such as when \(\textbf{M}\) models a predator-prey
system in which \(M_{ij}\) and \(M_{ji}\) have opposing signs, stability
increases\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}}. If
diagonal elements of \(\boldsymbol{\gamma}\) vary independently, the
magnitude of \(\rho\) is decreased because \(\sigma^{2}_{\gamma}\)
increases the variance of \(M_{ij}\) without affecting the expected
covariance between \(M_{ij}\) and \(M_{ji}\) (Figure 2).

@@

\textbf{Targeted manipulation of \(\boldsymbol{\gamma}\)}. To further
investigate the potential of \(\sigma^{2}_{\gamma}\) to be stabilising,
I used a genetic algorithm. Genetic algorithms are heuristic tools that
mimic evolution by natural selection, and are useful when the space of
potential solutions (in this case, possible combinations of \(\gamma\)
values leading to stability in a complex system) is too large to search
exhaustively\textsuperscript{\protect\hyperlink{ref-Hamblin2013}{29}}.
Generations of selection on \(\gamma\) value combinations to minimise
\(\Re(\lambda_{max})\) demonstrated the potential for
\(\sigma^{2}_{\gamma}\) to increase system stability. Across
\(S = \{2, 3, ..., 39, 40\}\), sets of \(\gamma\) values were found that
resulted in stable systems with probabilities that were up to four
orders of magnitude higher than when \(\gamma = 1\) (see Supplementary
Information), meaning that stability could often be achieved by
manipulating \(S\) \(\gamma\) values rather than \(S \times S\)
\(\mathbf{M}\) elements (i.e., by manipulating component response rates
rather than interactions between components).

@@

My focus here is distinct from Gibbs et
al.\textsuperscript{\protect\hyperlink{ref-Gibbs2017}{22}}, who applied
the same mathematical framework to investigate how a diagonal matrix
\(\mathbf{X}\) (equivalent to \(\boldsymbol{\gamma}\) in my model)
affects the stability of a community matrix \(\mathbf{M}\) given an
interaction matrix \(\mathbf{A}\) within a generalised Lotka-Volterra
model, where \(\mathbf{M} = \mathbf{XA}\). Gibbs et
al.\textsuperscript{\protect\hyperlink{ref-Gibbs2017}{22}} analytically
demonstrated that the effect of \(\mathbf{X}\) on system stability
decreases exponentially as system size becomes arbitrarily large
(\(S \to \infty\)) for a given magnitude of complexity
\(\sigma\sqrt{SC}\). My numerical results do not contradict this
prediction because I did not scale \(\sigma = 1 / \sqrt{S}\), but
instead fixed \(\sigma\) and increased \(S\) to thereby increase total
system complexity (see Supplemental Information for results simulated
across \(\sigma\) and \(C\)). Overall, I show that component response
rate variation increases the upper bound of complexity at which
stability can be realistically observed, meaning that highly complex
systems are more likely than not to vary in their component response
rates, and for this variation to be critical for system stability.

@@

A general mathematical framework encompassing shifts in eigenvalue
distributions caused by a diagonal matrix \(\boldsymbol{\gamma}\) has
been
investigated\textsuperscript{\protect\hyperlink{ref-Ahmadian2015}{21}}
and recently applied to questions concerning species density and
feasibility\textsuperscript{\protect\hyperlink{ref-Gibbs2017}{22},\protect\hyperlink{ref-Stone2017}{23}},
but \(\boldsymbol{\gamma}\) has not been interpreted as rates of
response of individual system components to perturbation. My model
focuses on component response rates for systems of a finite size, in
which complexity is high but not yet high enough to make the probability
of stability unrealistically low for actual empirical systems. For this
upper range of system size, randomly assembled complex systems are more
likely to be stable if their component response rates vary (e.g.,
\(10 < S < 30\) for parameter values in Fig. 4). Variation in component
response rate might therefore be critical for maintaining stability in
many highly complex empirical systems. These results are broadly
applicable for understanding the stability of complex networks across
the physical, life, and social sciences.

@@

\textbf{Component response rate (\(\boldsymbol{\gamma}\)) variation}. In
a synthesis of eco-evolutionary feedbacks on community stability, Patel
et al. model a system that includes a vector of potentially changing
species densities (\(\mathbf{n}\)) and a vector of potentially evolving
traits
(\(\mathbf{x}\))\textsuperscript{\protect\hyperlink{ref-Patel2018}{19}}.
For any species \(i\) or trait \(j\), change in species density
(\(n_{i}\)) or trait value (\(x_{j}\)) with time (\(t\)) is a function
of the vectors \(\mathbf{n}\) and \(\mathbf{x}\),

@@

\[\frac{d\mathbf{v}(t)}{dt} = \boldsymbol{\gamma} \mathbf{A}\mathbf{v}(t).\]

In the above, \(\boldsymbol{\gamma}\) is a diagonal matrix in which
elements correspond to individual component response rates. Therefore,
\(\mathbf{M} = \boldsymbol{\gamma} \mathbf{A}\) defines the change in
values of system components and can be analysed using the techniques of
May\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Ahmadian2015}{21},\protect\hyperlink{ref-May1973}{30}}.
In these analyses, row means of \(\mathbf{A}\) are expected to be
identical, but variation around this expectation will naturally arise
due to random sampling of \(\mathbf{A}\) off-diagonal elements and
finite \(S\). In simulations, the total variation in \(\mathbf{M}\) row
means that is attributable to \(\mathbf{A}\) is small relative to that
attributable to \(\boldsymbol{\gamma}\), especially at high \(S\).
Variation in \(\boldsymbol{\gamma}\) specifically isolates the effects
of differing component response rates, hence causing differences in
expected \(\mathbf{M}\) row means.

\textbf{Construction of random and structured networks}. I used the R
programming language for all numerical simulations and
analyses\textsuperscript{\protect\hyperlink{ref-Rproject}{31}}. Purely
random networks were generated by sampling off-diagonal elements from an
i.i.d \(A_{ij} \sim \mathcal{N}(0, 0.4)\) with a probability \(C\)
(unsampled elements were set to zero). Diagonal elements \(A_{ii}\) were
set to \(-1\). Elements of \(\boldsymbol{\gamma}\) were simulated i.i.d.
from a distribution with positive support (typically
\(\gamma_{i} \sim \mathcal{U}(0, 2)\)). Random \(\mathbf{A}\) matrices
with correlated elements \(A_{ij}\) and \(A_{ji}\) were built using
Cholesky decomposition. Competitor networks in which off-diagonal
elements \(A_{ij} \geq 0\) were constructed by first building a random
\(\mathbf{A}\), then flipping the sign of any elements in which
\(A_{ij} < 0\). Similarly, mutualist networks were constructed by
building a random \(\mathbf{A}\), then flipping the sign of elements
where \(A_{ij} > 0\). Predator-prey networks were constructed by first
buing a random \(\mathbf{A}\), then flipping the sign of either
\(A_{ij}\) or \(A_{ji}\) if \(A_{ij} \times A_{ji} > 0\).

@@

Cascade food webs were constructed following Solow and
Beet\textsuperscript{\protect\hyperlink{ref-Solow1998}{17}}. First, a
random matrix \(\mathbf{A}\) was generated with off-diagonal elements
sampled i.i.d so that \(A_{ij} \sim \mathcal{N}(0, 0.4)\). Each
component in the system was ranked from \(1\) to \(S\). If component
\(i\) had a higher rank than component \(j\) and \(A_{ij} < 0\), then
\(A_{ij}\) was multiplied by \(-1\). If \(i\) had a lower rank than
\(j\) and \(A_{ji} < 0\), then \(A_{ji}\) was multiplied by \(-1\). In
practice, this resulted in a matrix \(\mathbf{A}\) with negative and
positive values in the lower and upper triangles, respectively. Diagonal
elements of \(\mathbf{A}\) were set to \(-1\) and \(C = 1\). I simulated
cascade food webs across all combinations of
\(S = \{2, 3, ..., 59, 60\}\).

@@

\textbf{Genetic algorithm}. Ideally, to investigate the potential of
\(\sigma^{2}_{\gamma}\) for increasing the proportion of stable complex
systems, the search space of all possible \(diag(\boldsymbol{\gamma})\)
vectors would be evaluated for each unique \(\mathbf{M = \gamma A}\).
This is technically impossible because \(\gamma_{i}\) can take any real
value between 0-2, but even rounding \(\gamma_{i}\) to reasonable values
would result in a search space too large to practically explore. Under
these conditions, genetic algorithms are highly useful tools for finding
practical solutions by mimicking the process of biological
evolution\textsuperscript{\protect\hyperlink{ref-Hamblin2013}{29}}. In
this case, the practical solution is finding vectors of
\(diag(\boldsymbol{\gamma})\) that decrease the most positive real
eigenvalue of \(\mathbf{M}\). The genetic algorithm used achieves this
by initialising a large population of 1000 different potential
\(diag(\boldsymbol{\gamma})\) vectors and allowing this population to
evolve through a process of mutation, crossover (swaping \(\gamma_{i}\)
values between vectors), selection, and reproduction until either a
\(diag(\boldsymbol{\gamma})\) vector is found where all
\(\Re(\lambda) < 0\) or some ``giving up'' critiera is met.

For each \(S = \{2, 3, ..., 39, 40\}\), the genetic algorithm was run
for 100000 random \(\mathbf{M = \gamma A}\) (\(\sigma_{A} = 0.4\),
\(C = 1\)). The genetic algorithm was initialised with a population of
1000 different \(diag(\boldsymbol{\gamma})\) vectors with elements
sampled i.i.d from \(\gamma_{i} \sim \mathcal{U}(0, 2)\). Eigenanalysis
was performed on the \(\mathbf{M}\) resulting from each
\(\boldsymbol{\gamma}\), and the 20 \(diag(\boldsymbol{\gamma})\)
vectors resulting in \(\mathbf{M}\) with the lowest
\(\Re(\lambda_{max})\) each produced 50 clonal offspring with subsequent
random mutation and crossover between the resulting new generation of
1000 \(diag(\boldsymbol{\gamma})\) vectors. Mutation of each
\(\gamma_{i}\) in a \(diag(\boldsymbol{\gamma})\) vector occurred with a
probability of 0.2, resulting in a mutation effect of size
\(\mathcal{N}(0, 0.02)\) being added to generate the newly mutated
\(\gamma_{i}\) (any \(\gamma_{i}\) values that mutated below zero were
multiplied by \(-1\), and any values that mutated above 2 were set to
2). Crossover occurred between two sets of 100
\(diag(\boldsymbol{\gamma})\) vectors paired in each generation; vectors
were randomly sampled with replacement among but not within sets. Vector
pairs selected for crossover swapped all elements between and including
two \(\gamma_{i}\) randomly selected with replacement (this allowed for
reversal of vector element positions during crossover; e.g.,
\(\{\gamma_{4}, \gamma_{5}, \gamma_{6}, \gamma_{7}\} \to \{\gamma_{7}, \gamma_{6}, \gamma_{5}, \gamma_{4}\}\)
). The genetic algorithm terminated if a stable \(\mathbf{M}\) was
found, 20 generations occurred, or if the mean \(\boldsymbol{\gamma}\)
fitness increase between generations was less than 0.01 (where fitness
was defined as \(W_{\gamma} = -\Re(\lambda_{max})\) for \(\mathbf{M}\)).

@@

\textbf{Figure 1: Eigenvalue distributions of random complex systems.}
Each panel shows the real (x-axis) and imaginary (y-axis) parts of
\(S =\) 400 eigenvalues from random \(S \times S\) matrices.
(\(\textbf{a}\)) A system represented by a matrix \(\mathbf{A}\), in
which all elements are sampled from a normal distribution with
\(\mu = 0\) and \(\sigma_{A} = 1/\sqrt{S}\). Points are uniformly
distributed within the blue circle centred at the origin with a radius
of \(\sigma_{A} \sqrt{S} = 1\). (\(\textbf{b}\)) The same system as
\(\textbf{a}\) after including variation in the response rates of \(S\)
components, represented by the diagonal matrix \(\gamma\), such that
\(\mathbf{M} = \gamma\mathbf{A}\). Elements of \(\gamma\) are randomly
sampled from a uniform distribution from \(\min = 0\) to \(\max = 2\).
Eigenvalues of \(\mathbf{M}\) are then distributed non-uniformly within
the red circle centred at the origin with a radius of
\(\sqrt{\sigma^{2}_{A}(1 + \sigma^{2}_{\gamma})S} \approx\) 1.13.
(\(\textbf{c}\)) A different random system \(\mathbf{A}\) constructed
from the same parameters as in \(\textbf{a}\), except with diagonal
values of \(-1\). (\(\textbf{d}\)) The same system \(\textbf{c}\) after
including variation in component response rates, sampled from
\(\mathcal{U}(0, 2)\) as in \(\textbf{b}\).

@@

\textbf{Figure 4: Stability of large complex systems with and without
variation in component response rate (\(\boldsymbol{\gamma}\)).} The
\(\ln\) number of systems that are stable across different system sizes
(\(S = \{2, 3, ..., 49, 50 \}\)) given \(C = 1\), and the proportion of
systems for which variation in \(\gamma\) is critical for system
stability. For each \(S\), 1 million complex systems are randomly
generated. Stability of each complex system is tested given variation in
\(\gamma\) by randomly sampling \(\gamma \sim \mathcal{U}(0, 2)\).
Stability given \(\sigma^{2}_{\gamma}>0\) is then compared to stability
in an otherwise identical system in which
\(\gamma = E[\mathcal{U}(0, 2)]\) for all components. Blue and red bars
show the number of stable systems in the absence and presence of
\(\sigma^{2}_{\gamma}\), respectively. The black line shows the
proportion of systems that are stable when \(\sigma^{2}_{\gamma}>0\),
but would be unstable if \(\sigma^{2}_{\gamma}=0\) (i.e., the
conditional probability that \(\mathbf{A}\) is unstable given that
\(\mathbf{M}\) is stable).
